Deep Neural Architectures for Discourse Segmentation in E-Mail Based Behavioral Interventions

### 2018-11-26 paper-amia-summit-2019/manuscript.tex
initial final version by Alex

\noindent{\bf Abstract}
\textit{Communication science approaches to develop effective behavior interventions, such as motivational interviewing (MI), are limited by traditional qualitative coding of communication exchanges, a very resource-intensive and time-consuming process. This study focuses on the analysis of e-Coaching sessions, behavior interventions that are delivered via email and grounded in the principles of MI. A critical step towards automated qualitative coding of e-Coaching communication exchanges is segmentation of emails into fragments that correspond to MI behaviors. This study frames email segmentation task as a classification problem and utilizes word embeddings in conjunction with punctuation and part-of-speech features to address it. We experimentally evaluated conditional random fields (CRF), a traditional machine learning method, as well as deep learning methods, such as multilayer perceptron (MLP), bidirectional recurrent neural network (BRNN) and convolutional recurrent neural network (CRNN). Our results indicate that CRNN outperform CRF, MLP and BRNN for the task of email segmentation achieving 0.989 macro F1-score overall and 0.825 macro F1-score for new segment detection.}

@@

The limitation of this study is that e-Coaching data is collected from a single medical institute; formatting, style and email segment can be different in other settings. Therefore, there is a need to replicate the experiments in this study with different data sets. As future work, we are planning to evaluate the methods presented in this work on the datasets from other behavioral interventions.

@@

Segmentation is the first step of qualitative analysis of unstructured clinical communications, such as e-Coaching. Although several studies have focused on segmentation problem in biomedical context, they are limited to segmenting clinical text in EHR into sections and sentences. No previous studies considered the task of automated segmentation of clinical communications into groups of MI behaviors in the context of unstructured MI sessions. By comparing the performance of machine learning methods for the task of segmentation of e-Coaching emails, we found out that convolutional recurrent neural networks provides the best performance in terms of all performance metrics. Manual segmentation of e-Coaching data is a very resource-intensive and time-consuming task, which can significantly decrease the time and effort required to develop an effective behavioral intervention. Our proposed methods can help to identify textual segments corresponding to MI behaviors in unstructured clinical dialog, which can then be annotated with an MI behavior annotation model in a pipeline setting. Automated segmentation and annotation of e-Coaching emails can significantly decrease the time to identify effective communication strategies in e-mail based motivational interviewing.

\section*{Acknowledgments}
This study was supported by a grant from the National Institutes of Health, NIDDK R21DK108071, Carcone and Kotov, MPIs. We would like to thank the research staff and student assistants in the Department of Family Medicine and Public Health Sciences at Wayne State University School of Medicine for their help in preparing the training dataset.

>>>>

\noindent{\bf Abstract}
\textit{Communication science approaches to develop effective behavior interventions, such as motivational interviewing (MI), are limited by traditional qualitative coding of communication exchanges, a very resource-intensive and time-consuming process. This study focuses on the analysis of e-Coaching sessions, behavior interventions delivered via email and grounded in the principles of MI. A critical step towards automated qualitative coding of e-Coaching sessions is segmentation of emails into fragments that correspond to MI behaviors. This study frames email segmentation task as a classification problem and utilizes word and punctuation mark embeddings in conjunction with part-of-speech features to address it. We evaluated the performance of conditional random fields (CRF) as well as multi-layer perceptron (MLP), bi-directional recurrent neural network (BRNN) and convolutional recurrent neural network (CRNN) for the task of email segmentation. Our results indicate that CRNN outperforms CRF, MLP and BRNN achieving 0.989 macro F1-score overall and 0.825 macro F1-score for new segment detection.}

@@

%The limitation of this study is that e-Coaching data is collected from a single medical institute; formatting, style and email segment can be different in other settings. Therefore, there is a need to replicate the experiments in this study with different data sets. As future work, we are planning to evaluate the methods presented in this work on the datasets from other behavioral interventions.

@@

Segmentation is the first step of qualitative analysis of unstructured clinical communications, such as e-Coaching. Although several studies have focused on segmentation problem in biomedical context, they are limited to segmenting clinical text in EHR into sections and sentences. No previous studies considered the task of automated segmentation of clinical communications into groups of MI behaviors in the context of unstructured MI sessions. By comparing the performance of machine learning methods for the task of segmentation of e-Coaching emails, we found out that convolutional recurrent neural networks demonstrate the best performance in terms of all performance metrics. Manual segmentation of e-Coaching sessions is a very resource-intensive and time-consuming task, which can significantly decrease the time and effort required to develop an effective behavioral intervention. Our proposed methods can help to identify textual segments corresponding to MI behaviors in unstructured clinical dialog, which can then be annotated with an MI behavior annotation model in a pipeline setting. %Automated segmentation and annotation of e-Coaching emails can significantly decrease the time to identify effective communication strategies in e-mail based motivational interviewing.
As future work, we are planning to evaluate the methods presented in this work on the datasets from other behavioral interventions.

\section*{Acknowledgments}
This study was supported by a grant from the National Institutes of Health, NIDDK R21DK108071, Carcone and Kotov, MPIs. We would like to thank the student assistants in the Department of Family Medicine and Public Health Sciences at Wayne State University School of Medicine for their help in preparing the training dataset.
